### 2018-09-28 paper/paper.tex
Update abstract and discussions

% Please keep the abstract below 300 words
%\section*{Abstract}
%NEEDS TO BE EDITED...!!!
%Non-coding RNA molecules contribute to cellular function through diverse roles, including genome regulation, DNA and RNA repair, RNA splicing, catalysis, protein synthesis, and intracellular transportation  \cite{Morris2014-iu,Kung2013-se}. The mechanisms of these actions can only be fully understood in terms of native secondary and tertiary structures. When provided with a sufficient number of homologous sequences, the gold standard for secondary structure prediction continues to be comparative analysis  \cite{Gutell1992-hu}. Alternatively, the prevailing {\em computational} approach to secondary structure is through the Gibbs (thermal) equilibrium, by Monte Carlo sampling or approximating the minimum free energy (MFE) configuration \cite{Mathews1999-hc,Zuker1999-rc}. Aside from the necessary approximations, an enduring debate concerns the biological relevance of equilibrium configurations \cite{Levinthal1969-hi,Higgs2000-gj,Flamm2008-ir, Baker1994-px}. Here we adopt a kinetic perspective and argue that the existence of reliable folding on biologically relevant time scales suggests an {\em intra-molecular} statistical relationship between secondary and {\em primary} structures: as compared with other locations, nucleotide sequences in and around secondary-structure stems will have fewer Watson-Crick and wobble matches that are inconsistent with the native structure. An ``ambiguity index'', one for each pair of molecule and presumed secondary structure, measures the prevalence of false matches and hence the tendency to form metastable structures  incompatible with native structures. The ambiguity index statistically separates an ensemble of RNA molecules that operate as single entities (Group I and II Introns) from an ensemble that operates as protein-RNA complexes (SRP and tmRNAs), and ensembles of secondary structures determined by comparative analysis from ones based on thermal equilibrium. We find lower average ambiguity in single-entity RNA's than protein-RNA complexes, and, among single-entity RNA's, lower ambiguity with comparative analyses than equilibrium analyses. Both comparisons are supported by exact and highly significant hypothesis tests. These experiments, motivated by a hypothesized mechanism of folding, and the first of their kind, are consistent with folding to metastable but not necessarily equilibrium structures. 
%
%
%% Please keep the Author Summary between 150 and 200 words
%% Use first person. PLOS ONE authors please skip this step. 
%% Author Summary not valid for PLOS ONE submissions.   
%\section*{Author Summary}
%Recent discoveries indicate that, in addition to being a messenger between DNA and protein, RNA molecules assume a wide range of biological functions. For biological macromolecules, structure is function. Experimental determination of RNA structures is still time-consuming, and computational approaches are of great importance. The prevailing computational approach tries to find the structure with the minimum energy, yet the relevance of this minimum energy structure to the native structure is still hotly debated. In this paper, we adopt a kinetic perspective, and argue that more emphasis should be placed on the folding process when trying to develop computational methods for RNA structure prediction. We present some statistical analyses using the primary and secondary data (sequence and base-pairs data) of RNA molecules, based on the concept of ``local ambiguity", i.e. the molecule's tendency to ``make a mistake" at a certain location when forming secondary structures. Our results highlight deficiencies in the minimum-energy approach, and demonstrate the importance of considering kinetics as well as protein-RNA interactions in developing computational approaches for RNA secondary structure prediction.

@@

\section*{Discussion}

%From the above results, we can see that we verified a lot of the suspicions we had in the discussions of the exploratory analysis results. We demonstrated that we can indeed statistically seperate \textit{single-entity RNAs} from \textit{protein-RNA complexes}, using the ``ambiguity index". We can also statistically seperate the \textit{comparative analysis} structures from the \textit{minimum free energy} structures for \textit{single-entity RNAs}, using the ``ambiguity index", but we can't achieve this when we look at \textit{protein-RNA complexes}.
%
%These results suggest a potentially significant difference between the structure formation mechanisms on \textit{single-entity RNAs} and \textit{protein-RNA complexes}, and a qualitative difference between the \textit{comparative analysis} methods and the \textit{minimum free energy} methods when applied to \textit{single-entity RNAs}.
%
%Possible that MFE methods are way off, or that much better answers come from equilibrium samples or even centroids...
%
%in vivo structure is a compromise between stability and accessibility. 
%
%Our reasoning is mechanistic. 
%
%Given a favorable tertiary structure, it is likely that many nucleotide substitutions 
%This reasoning, mechanistic, 
%
%``Under Construction''!!!
%
%******
%The goal of this paper is to present statistical evidence that supports a a kinetic, but not necessarily equilibrium, viewpoint of RNA structure. verify our basic intuition, by looking at the differences in terms of local ambiguity between locations that participate in the secondary structure and the locations that don't, and to establish an intra-molecular statistical relationship between secondary and primary structures. This is achieved by some exploratory analysis as well as two sets of exact and highly significant hypothesis tests, which unveil fascinating results on the local ambiguities in different regions of the RNA molecules, the possible mechanistic differences in the structure formation of single-entity RNAs and protein-RNA complexes, as well as the subtle differences between the comparative analysis approach and the minimum free energy approach for RNA secondary structure prediction.
%**********
%
%Mention another possible statistical signature: stems that are phylogenetically older tend to be formed earlier in the folding process
%
%First, from the experiments and the results in this paper, ``local ambiguity" clearly emerges as a useful concept in the statistical analysis of RNA primary and secondary structure data and the folding process of RNA molecules.
%
%Based on ``local ambiguity", an ``ambiguity index'', one for each pair of molecule and presumed secondary structure, measures the prevalence of false matches and hence the tendency to form metastable structures  incompatible with native structures. The ambiguity index statistically separates an ensemble of RNA molecules that operate as single entities (Group I and II Introns) from an ensemble that operates as protein-RNA complexes (SRP and tmRNAs), and ensembles of secondary structures determined by comparative analysis from ones based on thermal equilibrium. We find lower average ambiguity in single-entity RNA's than protein-RNA complexes, and, among single-entity RNA's, lower ambiguity with comparative analyses than equilibrium analyses. These results demonstrate possible mechanistic differences in the structure formation of single-entity RNAs, where kinetics play a more important role, and protein-RNA complexes, where the protein-RNA interactions have significant impacts on the folding process, and a qualitative difference between the comparative analysis approach and the minimum free energy approach in the case where kinetics are important (in this paper, for single-entity RNAs).
%
%
%
%These empirical evidence points to the importance of carefully considering the impacts of kinetics and protein-RNA interactions on the folding process of non-coding RNA molecules, and argues against the naive application of thermal equilibrium based approaches for RNA secondary structure prediction.

>>>>

% Please keep the abstract below 300 words
\section*{Abstract}
When it comes to folding, the RNA energy landscape is particularly rugged, densely populated with 
energy-favorable pairings of nucleotide sequences, most of which are not a part of the molecule's native structure or structures. Yet many non-coding RNA molecules spontaneously fold to a native structure when placed in a suitable chemical environment at an appropriate temperature.
Levinthal's convincing arguments (aka Levinthal's paradox) about proteins apply to RNA molecules as well, perhaps even more compellingly: arrival to thermal equilibrium should take a great deal more time than the biological lifetime of the organism. Evidently, mechanisms have evolved that contribute to a directed and efficient folding process, and quite possibly the active configuration is not at all like a sample from an equilibrium distribution, but rather a metastable state, provided only that it will typically last long enough for its purpose. 
Various measures of the difficulty in reaching a stable configuration, whether or not it is a true equilibrium,
are based on the number of possible Watson-Crick or G$\cdot$U wobble pairings at each location. We call these measures ``ambiguity indexes,'' and focus on two, each of which is a function of both the primary and any given proposed secondary structure. Comparing families of RNA molecules that appear in ribonucleoproteins (``bound'' molecules) to RNA molecules that do not operate as nucleoproteins (``unbound'' molecules), we find that the unbound molecules have systematically lower ambiguity indexes, {\em provided that the indexes are computed using the consensus structure} (also known as the ``gold standard''). The effect largely disappears when the indexes are based, instead, on the so-called minimum-free energy (MFE) structures. Since MFE structures are derived from an approximation of the equilibrium distribution, we interpret these and other related results as evidence for metastable but not necessarily thermal equilibrium secondary structure. 
%
%
%% Please keep the Author Summary between 150 and 200 words
%% Use first person. PLOS ONE authors please skip this step. 
%% Author Summary not valid for PLOS ONE submissions.   
%\section*{Author Summary}
%Recent discoveries indicate that, in addition to being a messenger between DNA and protein, RNA molecules assume a wide range of biological functions. For biological macromolecules, structure is function. Experimental determination of RNA structures is still time-consuming, and computational approaches are of great importance. The prevailing computational approach tries to find the structure with the minimum energy, yet the relevance of this minimum energy structure to the native structure is still hotly debated. In this paper, we adopt a kinetic perspective, and argue that more emphasis should be placed on the folding process when trying to develop computational methods for RNA structure prediction. We present some statistical analyses using the primary and secondary data (sequence and base-pairs data) of RNA molecules, based on the concept of ``local ambiguity", i.e. the molecule's tendency to ``make a mistake" at a certain location when forming secondary structures. Our results highlight deficiencies in the minimum-energy approach, and demonstrate the importance of considering kinetics as well as protein-RNA interactions in developing computational approaches for RNA secondary structure prediction.

@@

\section*{Discussion}
We have introduced two indexes designed as coarse measures of the prevalence of kinetic traps in the energy landscape of a structural RNA molecule. These indexes, which we call the T-S and D-S ambiguity indexes, are each a function of both the primary and any presumed secondary structure of the molecule. The first, the T-S index, compares the availability of false pairings (i.e. stems that are not part of the purported secondary structure) of sequences straddling transitions from double to single-stranded regions to the availability of false pairings strictly within single-stranded regions.
The second, the D-S index, results from a similar calculation, but using double-stranded sequences in place of {\em transitional} sequences. We have given evidence that either index can be used effectively to classify unbound RNA (molecules that do not act as components of ribonucleoproteins) from bound RNA (those that are components of ribonucleoproteins), {\em provided that the secondary structure is derived from comparative analysis}. 
Specifically, unbound molecules typically have lower ambiguity indexes than bound molecules. On the other hand, there is little evidence for this effect in the ambiguity indexes derived instead from MFE analyses, and classifiers based on MFE secondary structure are substantially less accurate. Furthermore, either index can be used in unbound molecules to effectively classify a given secondary structure as derived from comparative as opposed to MFE analysis, but not in bound molecules. In particular, among unbound molecules comparative analysis structures typically have lower ambiguity indexes than MFE structures.

Both classifications, distinguishing unbound from bound molecules using comparative secondary structures and distinguishing comparative from MFE secondary structures in unbound molecules, are stronger when they are based on the T-S index rather than the D-S index. Though highly speculative, this would make sense if it were the case that stems were typically nucleated at one or the other end (what we have called {\em transitional} locations), as opposed to locations (sequences of four nucleotides) completely contained in a stem structure. In other words, does nucleation typically occur near the ends or in the interiors of stems? Our data appears to favor the former.

Among our goals is to stress the possible importance of kinetics, as opposed to thermal equilibrium {\em per se}, in the determination of native structure. Demonstrably, many non-coding RNA molecules will spontaneously fold and then remain in apparent equilibrium in their biologically active structures. 
But biological stability is relative to biological timescale; the folding of any particular RNA could just as well end in metastability, provided that the process is repeatable and the result sufficiently stable over the molecule's proper biological lifetime. Indeed, it would be arguably easier to evolve an effective tertiary structure without the additional and unnecessary burden of thermal equilibrium. 
However, and this has often been pointed out (e.g. xxx, xxx, xxx), the MFE structure itself may be a poor representative of thermal equilibrium. It is quite possible, then, that our observations to the effect that comparative and MFE structures have substantially different relationships to the ambiguity indexes, and our interpretation that comparative structures better separate unbound from bound molecules, would not hold up if we were to adopt a more ensemble-oriented structure in place of the MFE, as advocated by xxx or xxx, for example.

Our expectation was that {\em transitional} and {\em double} locations would be relatively more protected from kinetic traps than {\em single} locations, and hence the focus on D and T relative to S in the formulation of the indexes. But this coin has another side: high ambiguity of {\em single} locations could certainly discourage, if not actually stereochemically prevent, access to native stems. In a related vein, Lin et al. \cite{Lin2018-xl} have given evidence that competing stems which are inconsistent may {\em both} contain a high measure of information about the equilibrium distribution, suggesting that in such cases both forms could be active and the notion of {\em single} might itself be ambiguous. This analysis, though, does presume the relevance of the equilibrium distribution, against which Levinthal has made some compelling arguments.

Concerning bound molecules, some of the unpaired RNA sequences may be critical to function, as in the 
messenger RNA-like (``MLR'') region of tmRNA, and therefore relatively unambiguous. 
This effect would be expected to raise both the T-S and D-S indexes, especially if the molecule's secondary structure develops in conjunction with and dependent on the protein contributions to the ribonuclear complex. In sum, 
an operational role for unpaired nucleotides together with contextual effects on stem formations may help to explain the substantial drop in T-S and D-S indexes when going from unbound to bound molecules. More generally, 
there are many such lines of reasoning, most of which suggest new indexes that could be statistically explored, especially as the data bank of known structures and functions continues to grow.
